A replication of a simulation study on measuring and correcting bias of under-5 mortality estimation in populations affected by HIV/AIDS

### 2020-02-24 analysis/paper.Rmd
fix typos and add missing refs

## Model Description
The goal of this large-scale simulation used in the original paper is to build a wide range of demographic histories that accounts for the variability between actual populations, to measure how biases such as the independence assumption of mother and children survival rate in a population influenced by HIV/AIDS would exist. @quattrochi2019 used stochastic simulation methods to generate a set of population with input parameters such as demographic features, mortality rate for U5M, maternal mortality rate, HIV prevalence, etc. The input simulation parameters vary over a large interval for different age groups across different populations. In each time step of one year, each woman in the population has some probabilities to give birth for those who are 15 to 49 years old, being infected with HIV, initiating ART (if HIV-positive), and dying given the simulation parameters. Through the aging process in the simulation population, children born with infected HIV mothers will have a probability of getting infected while all children will have a probability of mortality.

For the calculation of bias, we used the formula provided in the original paper. 
$$\mbox{Relative bias} = \frac{IE_{survivors} - IE_{survivors,HIV deaths}}{IE_{survivors,HIV deaths}}$$

$$\mbox{Absolute bias} = IE_{survivors} - IE_{survivors,HIV deaths}$$
According to [@quattrochi2019], $IE_{survivor}$ was the estimates of U5M based on living women in 2010, and $IE_{survivors,HIV deaths}$ was the estimates of U5M based on living women in 2010 and women died from HIV/AIDS before 2010 given the condition that they would have been 15-49 in 2010 if they had survived. 
$a_{d}$

@@

Since the primary goal in @quattrochi2019 was to measure bias in indirect estimates across a set of populations that have experienced different rates of fertility, mortality, HIV infection, and ART (antiretroviral therapy) initiation. We adopted the same method to generate a set of population using the following then inputs: fertility, adult mortality, U5M, percent of 15–19 year olds who are sexually active, maternal mortality in 1990, percent annual decline in the maternal mortality rate, HIV incidence, duration of breastfeeding, and ART coverage. The original paper simulated one population for each combination of inputs resulting in a total of 4480 populations. Since the original paper did not mention the amount of time required to complete the simulation, we started our process with the same 4480 populations but without parallel computing, we could not complete the simulation within the timeframe of our study and stopped the simulation when it reached 1000. For the purpose of replication, we set the seed to be a different value before running the simulation.

@@

Initially, we attempted to change the year parameter and the population age parameter in the simulation construction process. We tried to change the year interval of our study from [1906, 2010] to [1906, 1966] as well as changing the population age interval from [0,120] to [0,75].  This turned out to be unsuccessful since the the above parameters were not only used in the simulation process but also in the following model selection and regression application. For instance, to estimate HIV prevalence, the model divides the population into 7 age groups, which cannot be achieved if the year or age parameter was reduced.
As a result, we used the same size and date for the initial population in the simulation. We initiated with 22,500 women who were aged 15 years in 1906, and ran the simulation through 2010.

@@

As mentioned in the original paper, according to [@chen2010fertility], women between 15–19 years old who were HIV-positive experienced higher ASFRs compared to HIV-negative women, depending if they were sexually active. Therefore the ASFR parameter was adjusted using the ratios estimated by [@chen2010fertility]. ASFR was also adjusted based on the studies that have found an increase in pregnancy following ART [@myer2010impact][@makumbi2011associations][@homsy2009reproductive]. The effect of ART on fertility likely depends on a variety of factors including age, cluster of differentiation 4 (CD4) count at initiation according to the original paper. For the simulation model, the assumption that the ASFR for 15–19 year olds is not affected by ART was made to further simplify the simulation.
Inputs relating to maternal mortality was calculated using MMR (maternal deaths per 100,000 births) in 1990 and the annual decline in MMR since 1990. The initial value of the MMR according to [@hogan2010maternal] was either 0.0012 or 0.012 and the annual rate of decline was set to 0 or 7.3% and calculated as:

@@

Newly-generated simulation data was then imported and fitted to various models to select for a best-performing model which provides least errors.

The simulated data was first divided into training set (80%) and test set (20%). The training set was fitted to eight different models, which were further evaluated by calculating several error measures (root mean square error, root median square error, mean relative error and median relative error) on both training set (in-sample) and test set (out-of-sample) shown in Table 4.

The glmnet with alpha = 1 was selected as the best-performing model based on its error measures and was used in predictions later.

@@

The figure 4 (Fig. \@ref(fig:ori-fig-4)) and 5 (Fig. \@ref(fig:ori-fig-5)) in the original article were created by applying the best-performing model as well as an independent model [@ward2008effect] to empirical data from 2010 DHS for Malawi and Tanzania respectively to estimate U5M of different age groups of population. By comparing differences in U5M estimates of unadjusted data and adjusted data by two models, the authors suggested that the new model was able to correct the bias from indirect estimates.

@@

To prepare data for plotting, we created three functions with R code provided by the authors to estimate U5M by indirect method, Ward-Zaba model and our best-performing model respectively. These three functions were then further integrated into a function called plot_fig to create our replicated figures. We also included a file with necessary data for calculating the confidence interval by bootstrapping for prediction with our best-fitting model, since this step usually takes more than 40 minutes to finish.

@@

We revised some parts of the regression model and found inconsistent variable names during the debugging process for the reproduction of the tables and figures to be sucessful. In the following section of the original regression code, we changed the parameter in the bootstrapping model from 1:500 to 1:5 to make it run faster. We also caught the inconsistency in the variable  “lambda.min” where the parameter should be “lambda” from the model.

@@

Table 1 (Fig. \@ref(fig:table1)) compares the prediction errors across the 13 models in the same format as in Table 4 in the original paper. Compared to the estimation errors calculated in the original paper, we had very similar results of the root mean square errors and root median square errors for all models. However, our in-sample mean relative errors were consistently greater compared to the original paper and we suspect this could be due to the reduction in population size during the simulation. Consistent with the original paper, no single model in the table showed clear advantages for all the metrics over others we selected the generalized linear regression with alpha equal to 1 as our predictive model since it showed the best performance for most of the metrics.

@@

# Conclusion
Our goal of this project is to replicate a published research project and allow readers to produce identical figures to those in the target paper. We were able to accomplish this goal by replicating the key figures in the original paper using R and publishing well-structured source code along with data source in a licensed Github repository. This paper.md R Markdown file will be used as the main entry point of our project and is fully executable. However, we intentionally set a couple code chunks with evaluation as False since simulating those data sources does not fit within the timeframe for any group who is trying to reproduce them. Instead, we created .RData file as a battery-included solution for future reproducers.

We leveraged some useful Reproducibility techniques introduced in the class such as Tidyverse libraries and R Markdown syntax. The in-file executable R chunks come extremely handy during our reproducible process.

There are, however, challenges throughout our explorations. The first major issue we encountered was the lengthy source data generation step which required extensive research and deep dive into source code. Given the nature of the HIV/AIDS estimation research, the data samples are required to be large enough and comprehensive which should cover across multiple geological locations and a full range of populations. Therefore, the original authors designed the raw data to be generated on-the-go rather than archived in open source. Our team underestimated the challenge of reproducing all sources by the time we picked this article. To tackle this, all team members gathered together and agreed to make tradeoffs - reducing population size during simulation. The decision was made in favor of the balance between quality of our research and a timely deliverable. We identified this as an area to improve (focus on verification of data source before replication) during our retrospective of this project. Another challenge was using R in an unfamiliar field of research. Given none of us had previous domain knowledge regarding HIV/AIDS research, we spent a lot of time understanding the terminologies while exploring the rationale behind the R code used in the original paper. We had multiple issues trying to reproduce a specific failure in R from multiple environments. For example, we still couldn’t fully understand some of the issues during “knit” failures which had been resolved by cleaning up cache in R Studio. We strongly believe that using Docker or a shared development environment could dramatically improve our developer experience where all bugs could be reproduced easily no matter where the Docker image is running.

In general, we gained hands-on experience from reproducing an entire research project. After rounds of reviews and merging conflicts, we developed our own Github usage convention and code styles. We would like to carry these invaluable skills to our future Data Science journey.

>>>>

## Model Description
The goal of the large-scale simulation used in the original paper was to build a demographic history that accounted for the variability among actual populations to measure if biases such as the independence assumption of mother and children survival rate in a population influenced by HIV/AIDS existed. @quattrochi2019 used stochastic simulation methods to generate a set of populations with input parameters including demographic features, mortality rate for U5M, maternal mortality rate, HIV prevalence, etc. The input simulation parameters vary over a large interval for different age groups across different populations. In each time step of one year, each woman in a population has some probabilities to give birth (for women between 15 to 49 years old), being infected with HIV, initiating ART (if HIV-positive), and dying given the simulation parameters. Through the aging process in the simulation population, children born with infected HIV mothers will have a probability of getting infected while all children will have a probability of mortality.

For calculating relative and absolute bias, we used the formulas provided in the original paper. 
$$\mbox{Relative bias} = \frac{IE_{survivors} - IE_{survivors,HIV deaths}}{IE_{survivors,HIV deaths}}$$

$$\mbox{Absolute bias} = IE_{survivors} - IE_{survivors,HIV deaths}$$
According to [@quattrochi2019], $IE_{survivor}$ was the estimattion of U5M based on living women in 2010, and $IE_{survivors,HIV deaths}$ was the estimation of U5M based on living women in 2010 and women died from HIV/AIDS before 2010 given the condition that they would have been 15-49 in 2010 if they had survived. 
$a_{d}$

@@

Since the primary goal in @quattrochi2019 was to measure bias in indirect estimates across a set of populations that have experienced different rates of fertility, mortality, HIV infection, and ART (antiretroviral therapy) initiation. We adopted the same method to generate a set of populations using the following inputs: fertility, adult mortality, U5M, percent of 15–19 year olds who are sexually active, maternal mortality in 1990, percent annual decline in the maternal mortality rate, HIV incidence, duration of breastfeeding, and ART coverage. The original paper simulated one population for each combination of inputs resulting in a total of 4480 populations. Since the original paper did not provide information on the amount of time required to complete the simulation, we started our process with the same 4480 populations but without parallel computing, we could not complete the simulation within the timeframe of our study and stopped the simulation when it reached 1000 populations. For the purpose of our replication study, we set the seed for simulation to be 1, which was different from the seed value 1000 in the original paper.

@@

Initially, we attempted to some parameters in the simulation construction process. We tried to reduce the year interval of our study from [1906, 2010] to [1906, 1966] as well as the population age interval from [0,120] to [0,75]. This turned out to be unsuccessful since the the above parameters were not only used in the simulation process but also in the following model selection and regression application. For instance, to estimate HIV prevalence, the model divides the population into 7 age groups, which cannot be achieved if the year or age parameter was reduced. As a result, we used the same size and date for the initial population in the simulation. We initiated with 22,500 women who were aged 15 years in 1906, and ran the simulation through 2010.

@@

As mentioned in the original paper, according to [@chen2010fertility], women between 15–19 years old who were HIV-positive experienced higher ASFRs compared to HIV-negative women, depending if they were sexually active. Therefore the ASFR parameter was adjusted using the ratios estimated by [@chen2010fertility]. ASFR was also adjusted based on the studies that have found an increase in pregnancy following ART [@myer2010impact][@makumbi2011associations][@homsy2009reproductive]. The effect of ART on fertility likely depends on a variety of factors including age, cluster of differentiation 4 (CD4) count at initiation according to the original paper. For the simulation model, the assumption that the ASFR for 15–19 year olds is not affected by ART was made to further simplify the simulation.
Inputs relating to maternal mortality was calculated using MMR (maternal deaths per 100,000 births) in 1990 and the annual decline in MMR since 1990. The initial value of the MMR according to [@hogan2010maternal] was either 0.0012 or 0.012 and the annual rate of decline was set to 0 or 7.3%.

@@

Newly-generated simulation data were then imported and fitted to various models to select for a best-performing model and the selection process used least errors to represent model performances.

The simulated data were first divided into training set (80%) and test set (20%). The training set was fitted using eight different models, which were further evaluated by calculating several error metrics (root mean square error, root median square error, mean relative error and median relative error) on both training set (in-sample) and test set (out-of-sample) shown in Table 1(Fig. \@ref(fig:table1)).

Compared to the estimation errors calculated in the original paper, we had very similar results of the root mean square errors and root median square errors for all models. However, our in-sample mean relative errors were consistently greater compared to the original paper and we suspect this could be due to the reduction in population size during the simulation. Consistent with the original paper, no single model in the table showed clear advantages for all the metrics over others we selected the generalized linear regression with alpha equal to 1 as our predictive model since it showed the best performance for most of the metrics.

@@

Figure 4 (Fig. \@ref(fig:ori-fig-4)) and 5 (Fig. \@ref(fig:ori-fig-5)) in the original article were created by applying the best-performing model as well as an independent model [@ward2008effect] to the empirical data from 2010 DHS for Malawi and Tanzania [@nationalbureauofstatistics-nbs_tanzania_icf_macro_2011][@nationalstatisticaloffice_icf_macro_2011] respectively to estimate U5M of different age groups of population. By comparing differences in U5M estimates of unadjusted data and adjusted data by two models, the authors suggested that the new model was able to correct the bias from indirect estimates.

@@

To prepare data for plotting, we created three functions with R code provided by the authors to estimate U5M using the indirect method, the Ward-Zaba model and our best-performing model. These three functions were then further integrated into a function called plot_fig to create our replicated figures. We also included a file with necessary data for calculating the confidence interval by bootstrapping for prediction with our best-fitting model, since this step usually takes more than 40 minutes to finish.

@@

We revised some parts of the regression model and corrected inconsistent variable names in the original code during the debugging process. For instance, caught corrected the variable name “lambda.min” where it should have been “lambda”.

@@

@@

# Conclusion
Our goal of this project was to replicate a published research paper and reproduce figures in it. We were able to accomplish this goal by replicating the key figures in the original paper using R and publishing well-structured source code along with data source in a licensed Github repository. This paper.Rmd R Markdown file is fully executable therefore can be used as the main entry point to our project. However, we intentionally set a couple code chunks with evaluation as False since simulating those data sources does not fit within the timeframe for any group trying to reproduce them. Instead, we created .RData file as a battery-included solution for future reproducers.

We leveraged some useful reproducibility techniques introduced in the class such as Tidyverse libraries and R Markdown syntax. The in-file executable R chunks come extremely handy during our reproduction process.

There are, however, challenges throughout our explorations. The first major issue we encountered was the lengthy source data generation step which required extensive research and deep dive into the source code. Given the nature of the HIV/AIDS estimation research, the data samples required had to be comprehensive enough to cover multiple geological locations and a wide range of populations. Therefore, the original authors designed the raw data to be generated on-the-go rather than archived in open source. Our team initially underestimated the challenge of reproducing all sources and this was also due to the lack of information on completion time in the original paper. We had to reduce population size during simulation to ensure the quality of our research and a timely deliverable. Looking back, we can identify this as an area that we could improve (focusing on verification of data source before replication). Another challenge was using R in an unfamiliar field of research. Given none of us had previous domain knowledge regarding HIV/AIDS research, we spent a lot of time to understand the terminologies while exploring the rationales behind the R code used in the original paper. We had issues trying to understnad specific failures in R givern that multiple team members had different R environment. For example, we still couldn’t fully understand some of the “knit” failures which had been resolved by cleaning up cache in R Studio. We strongly believe that using Docker or a shared development environment could drastically improve our developer experience where most errors could be reproduced easily no matter where the Docker image is running.

In general, we gained hands-on experience from reproducing a research project. After rounds of reviews and merging conflicts, we developed our own Github usage convention and code styles. We would like to carry these invaluable skills to our future Data Science journey.
